Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-07-25 chap-intro.tex
Collect excluded papers, emphasize thread about modularity
Close #15.

What's worse, some of these manual optimizations are best done over the whole
program. For instance, adding an index on some collection can speed up looking
for some information, but each index must be maintained incrementally when the
underlying data changes. Depending on the actual queries and updates performed
on the collection, and on how often they happen, it might turn out that updating
an index takes more time than the index saves; hence the choice of which indexes
to enable depends on the whole program. However, adding/removing an index
requires updating all PL queries to use it, while RDBMS queries can use an index
transparently.

\section{This thesis}
To reduce the need for manual optimizations, in this thesis we propose
techniques for optimizing higher-order collection queries and executing them
incrementally.
\pg{WIP: What's my thesis? ``We provide approaches'' is better but not yet enough.}
By generalizing database-inspired approaches, we provide approaches to query optimization and
incrementalization by code transformation to apply to higher-order collection
queries including user-defined functions.
%While
\pg{Status: others have done much more on optimization.}
Instead of building on existing approaches to incrementalization, such as
self-adjusting computation~\citep{Acar09}, we introduce a novel
incrementalization approach which enables further transformations on the
incrementalization results.
%
Further research is needed, but a number of case studies suggest
applicability, even in some scenarios beyond existing
techniques~\citep{Koch2016incremental}.

We consider the problem for functional programming languages such as Haskell or
Scala, and we consider collection queries written using the APIs of their
collection libraries, which we treat as an embedded domain-specific language (EDSL). Such
APIs (or DSLs) contain powerful operators on collections such as $\Varid{map}$,\pg{exemplify or take for granted?}
which are higher-order, that
is they take as arguments arbitrary functions in the host language that we must
also handle.

@@

\pg{Readd section on excluded papers.}
% \subsection{Excluded papers}
% During my PhD work I collaborated on several other papers.

% \paragraph{Software Product Lines}
% Initial work on Software Product Lines lead to \cite{KPO:VaMoS11} and
% \cite{KGREOB:OOPSLA11}. On the same line I collaborated to experiments in
% \cite{SRKGAK:SPLC11} and its updated version\pg{TODO}.
% % \pg{Cite work on VarPL? Unlikely.}

% \paragraph{Modularity}
% \cite{OGKR:ECOOP11}

% \paragraph{DSL work}
% While working on modular embedded DSLs I collaborated to \citet{ErdwegGR12},
% \pg{TODO GADT paper}, \pg{TODO maybe manuscript on cake pattern?}.

>>>>

What's worse, some of these manual optimizations are best done over the whole
program. Some optimizations only become possible after inlining, which breaks
modularity if done by hand.
Worse, adding an index on some collection can speed up looking
for some information, but each index must be maintained incrementally when the
underlying data changes. Depending on the actual queries and updates performed
on the collection, and on how often they happen, it might turn out that updating
an index takes more time than the index saves; hence the choice of which indexes
to enable depends on the whole program. However, adding/removing an index
requires updating all PL queries to use it, while RDBMS queries can use an index
transparently.

Overall, manual optimizations are not only effort-intensive and error-prone, but
they also significantly reduce modularity.

\section{This thesis}
To reduce the need for manual optimizations, in this thesis we propose
techniques for optimizing higher-order collection queries and executing them
incrementally.
\pg{WIP: What's my thesis? ``We provide approaches'' is better but not yet enough.}
By generalizing database-inspired approaches, we provide approaches to query optimization and
incrementalization by code transformation to apply to higher-order collection
queries including user-defined functions.
%While
\pg{Status: others have done much more on optimization.}
Instead of building on existing approaches to incrementalization, such as
self-adjusting computation~\citep{Acar09}, we introduce a novel
incrementalization approach which enables further transformations on the
incrementalization results. This approach is in fact not restricted to
collection queries but applicable to other domains, but it requires adaptation
for the domain under consideration.
%
Further research is needed, but a number of case studies suggest
applicability, even in some scenarios beyond existing
techniques~\citep{Koch2016incremental}.

We consider the problem for functional programming languages such as Haskell or
Scala, and we consider collection queries written using the APIs of their
collection libraries, which we treat as an embedded domain-specific language
(EDSL). Such APIs (or DSLs) contain powerful operators on collections such as
$\Varid{map}$,
which are higher-order, that
is they take as arguments arbitrary functions in the host language that we must
also handle.

@@

\pg{Readd section on excluded papers.}

\subsection{Excluded papers}
This thesis improves modularity of collection queries by automating different sorts of
optimizations on them.

During my PhD work I collaborated on several other papers, mostly on different facets of
modularity, which do not appear in this thesis.

\paragraph{Modularity}
Module boundaries hide information on implementation that is not relevant to
clients. However, it often turns out that some scenarios, clients or task
require such hidden information. As discussed, manual optimization requires
hidden information: hence, researchers strive to automate optimizations. But
other tasks often violate these boundaries. I collaborated on research on
understanding why this happens and how to deal with this
problem~\citep*{OGKR:ECOOP11}.

\paragraph{Software Product Lines}
In some scenarios, a different form of modular software is realized through
software product lines (SPLs), where many \emph{variants} of a single software
artifact (with different sets of features) can be generated. This is often done
using conditional compilation, for instance through the C preprocessor.

But if a software product line uses conditional compilation, processing
automatically its source code is difficult---in fact, even lexing or parsing it
is a challenge. Since the number of variants is exponential in the number of
features, generating and processing each variant does not scale.

To address these
challenges, I collaborated with the TypeChef project to develop a
variability-aware lexer \citep*{KPO:VaMoS11} and parser
\citep*{KGREOB:OOPSLA11}.

Another challenge is predicting non-functional properties (such as code size or
performance) of a variant of a software product line before generation and
direct measurement. Such predictions can be useful to select efficiently a
variant that satisfies some non-functional requirements.
I collaborated to research on predicting such non-functional properties: we
measure the properties on a few variants and extrapolate the results to other
variants. In particular, I collaborated to the experimental evaluation on a few
open source projects, where even a simple model reached significant accuracy in a
few scenarios \citep*{SRKGAK:SPLC11,siegmund2013scalable}.

\paragraph{Domain-specific languages}
This thesis is concerned with DSLs, both ones for the domain of collection
queries, but also (in \cref{part:incr}) more general ones.

Different forms of language composition, both among DSLs and across
general-purpose languages and DSLs are possible, but their relationship is
nontrivial; I collaborated to work classifying the different forms of language
compositions~\citep*{ErdwegGR12}.

The implementation of \LoS, as described in \cref{part:ch-aosd13}, requires an
extensible representation of ASTs, similar to the one used by LMS
\citep{rompf2010lightweight}. While this representation is pretty flexible, it
relies on Scala's support of GADTs \citep{Emir06Variance,Emir07Patterns}, which
is known to be somewhat fragile due to implementation bugs. In fact,
sound pattern matching on Scala's extensible GADTs is impossible
without imposing significant restrictions to extensibility, due to language
extensions not considered during formal study
\citep{Emir06Variance,Emir07Correct}: the problem arises due to the interaction
between GADTs, declaration-site variance annotations and variant refinement of
type arguments at inheritance time.
To illustrate the problem, I've shown it already applies to an extensible
definitional interpreter for $\lambda_{<:}$~\citep{Giarrusso2013gadt}.
While solutions have been investigated in other settings
\citep{Scherer13GADTSub}, the problem remains open for Scala to this day.

%\pg{TODO maybe manuscript on cake pattern?}.
